---
title: "Spatial Ecology Lab Exercise 1"
author: "Kyle Falk"
date: "Fall 2025"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

#Setting up Working Directory
```{r}

library(ggplot2)
library(here)
library(dplyr)

getwd()

#install packages if not already installed
if(!require(terra)) install.packages("terra")
if(!require(dplyr)) install.packages("dplyr")
if(!require(ggplot2)) install.packages("ggplot2")

#load required packages
library(terra) #load 'terra' for raster/spatial analysis
library(dplyr) #load 'dplyr' for data manipulation
library(ggplot2) #load 'ggplot2' for plotting
library(ggridges) #load 'ggridges' for ridgeline plots (good for visualizing distributions, e.g., DOY patterns)
library(purrr) #load 'purrr' for functional programming tools (map, map_dfr, etc. useful for batch processing)

```


#Load Raster
``` {r}

#let's try to read the data in 2022

getwd()
setwd("/home/guest/ENV 567/Spatial_Fall2025/Lab Exercises/h027v011/h027v011")

#read the disturbance agent raster. Each pixel value represents the type of disturbance agent.
agent_2022 <- rast("./CD_027011_2022_C01/CD_027011_2022_C01_AGENT.tif")

#read the disturbance severity raster. Each pixel value indicates the severity level of the disturbance.
severity_2022 <- rast("./CD_027011_2022_C01/CD_027011_2022_C01_SEVERITY.tif")

#read the disturbance time raster. Each pixel value represents the day of year of disturbance occurrence.
time_2022 <- rast("./CD_027011_2022_C01/CD_027011_2022_C01_TIME.tif")




